
# Pre-processing of raw SORTseq files (QC, filtering)
Written by: Peter Brazda (20 Feb 2024) - Last Update: Farid Keramati, Peter Brazda (13 March 2024)

This workflow aims to help an initial, per-plate (visual) inspection (1) and to prepare expression matrices for the downstream steps (2):

QC
1.1 Set the environment
1.2 Collect the input files
1.3 Run the loop to generate the .pdf reports

Filtering
2.1 Set the environment
2.2 Collect the input files
2.3 Run the loop to generate the filtered matrices

Warning: Before running the script make sure to download Functions_QC.R for the helper functions, hg38_genes_with_ERCC_unique_ensembl_gene.tab for the genome annotation, and Plate_distribution_simple.xlsx for the 384-well plate format and the raw matrix files (e.g., TM621.transcripts). The format of the name is important.
The pipeline requires an output folder for the compiled .pdf reports and also a 'temp' folder for the intermediate .png files.

## 1. Generate QC-reports per plate

### 1.1 Set the environment
Identify the input/output folders and load the helper files and install the libraries

```{r}

home <- "C:/Users/Malieva/Desktop/napo/BGT-dev_avi/scripts" # select home directory

setwd(home)

source(paste0(home,"/helper_files/Functions_QC.R")) # the function to generate the QC plots for one plate
dir <-  paste0(home,"/out/qc_plots/") # location for the final output .pdf file (the per-plate visual report)
temp <- paste0(home,"/out/qc_plots/temp") # location for the temporary .png files
matrices <- paste0(home,"/in/") # location for the raw matrix files
gencode <- read.table(paste0(home,"/helper_files/hg38_genes_with_ERCC_unique_ensembl_gene.tab"), header = TRUE) # the hg38 reference extended with the ERCC genes
primer <- read_excel(paste0(home,"/helper_files/Plate_distribution_simple.xlsx"), col_names = FALSE) # primer plate distribution file
```

```{r}
ls()
```

### 1.2 Collect a list of all .transcript.txt files

```{r}
transcript_files <- list.files(matrices, '\\.transcripts.txt$', full.names = TRUE)
```

### 1.3 Loop through each file to generate QC report
It is advised to remove the suppressWarnings during the first run.

```{r}
suppressWarnings({

    for(file_path in transcript_files) {
  setwd(home)
  plateID <- gsub("\\.transcripts\\.txt$", "", basename(file_path)) # extract the plateID from the file name
  setwd(temp)
  
  # Generate QC report for the current plate
  Prepare_SORTseq_QC_report(plateID = plateID, matrices = matrices, primer = primer, gencode = gencode) # the function is defined in Functions_QC.R
  
  # Grab the files from the temp folder
  files <- list.files(temp, '\\.png$', full.names = TRUE)
  rl <- lapply(files, png::readPNG)
  gl <- lapply(rl, grid::rasterGrob)
  
  pdf_file <- paste0(dir, "QCreport_", plateID, ".pdf")
  ggsave(pdf_file, gridExtra::marrangeGrob(grobs = gl, nrow = 5, ncol = 4, top = NULL), width = 8, height = 10)
  file.remove(files) # Remove temporary .png files after saving the report
}
    })
    
```

## 2. Filter the matrices

### 2.1 Set the environment
Identify the input/output folders and load the helper files.

```{r}
home <- "path to your folder" # select home directory


setwd(home)
source(paste0(home,"/helper_files/Functions_QC.R")) # the function to generate the QC plots for one plate
dir <-  paste0(home,"/out/qc_plots/") # location for the final output .pdf file (the per-plate visual report)

gencode <- read.table(paste0(home,"/helper_files/hg38_genes_with_ERCC_unique_ensembl_gene.tab"), header = TRUE) # the hg38 reference extended with the ERCC genes
input_location <- paste0(home,"/in/") # location for the raw matrix files
output_location_nME <- paste0(home,"/out/qc_plots/1_noMITO_noERCC/") # output folder for the matrix after filtering and subsequent removal of mitochondrial AND ERCC genes
output_location_nE <- paste0(home,"/out/qc_plots/2_noERCC/")  # output folder for the matrix after filtering and subsequent removal of ERCC genes

```

### 2.2 Collect a list of all .transcript.txt files

```{r}
transcript_files <- list.files(input_location, pattern = "\\.transcripts\\.txt$", full.names = FALSE)
```

### 2.3 Loop through each file to process them
This loop uses a custom function that identifies mitochondrial and ERCC genes and generates matrices without these genes.
It is advised to remove the suppressWarnings during the first run.

```{r}
suppressWarnings({
    for(file_name in transcript_files) {
  plateID <- gsub("\\.transcripts\\.txt$", "", file_name) # extract the plateID from the file name
  
  # Call the filtering function for each plate
  Filter_SORTseq_tables(plate = plateID,  # the function is defined in Functions_QC.R
                        input_location = input_location,
                        # mito.pct.max = mito.pct.max, # set the upper limit for % mitochondrial gene counts per cell, default = 30
                        # ercc.content.max = ercc.content.max # set the upper limit for % ERCC gene counts per cell (high % reflects failed mRNA-capture, but successful amplification), default = 25
                        output_location_nME = output_location_nME,
                        output_location_nE = output_location_nE) 
}
    })
```

```{r}
print(sessionInfo())
```

